---
title: "NHL Goalies Capstone Project- Data Cleaning"
output: word_document
date: "2026-02-27"
author: "Lana Robison"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data Joining & Cleaning


## NHL Edge Goalie Data: Data Cleaning

- Join Data from NHL Edge Vertically

```{r}
library(readxl)
library(tidyverse)
library(janitor)
library(stringr)

nhl_edge_path <- "C:/Users/lanac/OneDrive - Baylor University/STA 43C8 Capstone/NHL Edge Goalie Data"

nhl_files <- list.files(nhl_edge_path, 
                        pattern = "\\.xlsx$", 
                        full.names = TRUE)

nhl_edge_raw_all <- map_dfr(
  nhl_files,
  ~ read_excel(.x, col_types = "text") %>%
    clean_names() %>%
    mutate(
      # keys 
      season = as.numeric(season),

      # numeric columns 
      gp = as.numeric(gp),
      gs = as.numeric(gs),
      cg = as.numeric(cg),
      ig = as.numeric(ig),
      cg_percent = as.numeric(cg_percent),
      gs_900 = as.numeric(gs_900),
      gs_900_percent = as.numeric(gs_900_percent),

      gf = as.numeric(gf),
      ga = as.numeric(ga),
      gfa = as.numeric(gfa),
      gaa = as.numeric(gaa),

      row = as.numeric(row),
      rol = as.numeric(rol),
      sa_60 = as.numeric(sa_60),
      sv_percent = as.numeric(sv_percent)
    ),
  .id = "file_id"
)


```



```{r}
nhl_edge_clean <- nhl_edge_raw_all %>%
  mutate(
    player = str_squish(player),
    team   = str_to_upper(str_squish(team)),
    season = floor(season / 10000)
  ) %>%
  filter(!is.na(player), player != "", !is.na(team), team != "") %>%
  distinct(player, team, season, .keep_all = TRUE)


```
- important to note that while seasons are technically labelled as 2008-2009, all seasons in this project will just be referred to by the starting years, so in this instance 2008-2009 season is the 2008 season




## Moneypuck Goalie & Team Data (2008-09 to 2024-25 season): Data Cleaning

Unzip files
```{r}
mp_zip_goalies <- "C:/Users/lanac/OneDrive - Baylor University/STA 43C8 Capstone/goalies_2008_to_2024.zip"

mp_zip_teams   <- "C:/Users/lanac/OneDrive - Baylor University/STA 43C8 Capstone/teams_2008_to_2024.zip"

dir.create("data_raw", showWarnings = FALSE)

unzip(mp_zip_goalies, exdir = "data_raw")
unzip(mp_zip_teams,   exdir = "data_raw")

list.files("data_raw")


```


```{r}
mp_goalies_raw <- read_csv("data_raw/goalies_2008_to_2024.csv") %>%
  clean_names()

mp_teams_raw   <- read_csv("data_raw/teams_2008_to_2024.csv") %>%
  clean_names()


```


```{r}
mp_goalies_clean <- mp_goalies_raw %>%
  mutate(
    player    = str_squish(name),
    team      = str_to_upper(str_squish(team)),
    season    = as.numeric(season),
    situation = str_to_lower(str_squish(situation)),
    position  = str_to_upper(str_squish(position))
  ) %>%
  filter(position == "G", situation == "all")

mp_teams_clean <- mp_teams_raw %>%
  mutate(
    team      = str_to_upper(str_squish(team)),
    season    = as.numeric(season),
    situation = str_to_lower(str_squish(situation))
  ) %>%
  filter(situation == "all")

```


Check distinct teams
```{r}
sort(unique(mp_goalies_clean$team))
sort(unique(mp_teams_clean$team))
sort(unique(nhl_edge_clean$team))
```

Team Name Standardization
```{r}
standardize_team <- function(x) {
  x <- str_squish(x)
  x <- str_to_upper(x)
  x <- str_replace_all(x, "\\.", "")
  
  x <- recode(
    x,
    "PHX" = "ARI",
    "LA"  = "LAK",
    "NJ"  = "NJD",
    "SJ"  = "SJS",
    "TB"  = "TBL",
    .default = x
  )
  
  return(x)
}

```


Apply function to moneypuck data
```{r}
mp_goalies_clean <- mp_goalies_clean %>%
  mutate(team = standardize_team(team))

mp_teams_clean <- mp_teams_clean %>%
  mutate(team = standardize_team(team))
```



keep most recent team in NHL edge data
```{r}
nhl_edge_clean <- nhl_edge_clean %>%
  mutate(
    team = str_squish(team),
    team = if_else(str_detect(team, ","), word(team, -1, sep = ","), team),
    team = standardize_team(team)
  )
```


verify team codes are standardized 
```{r}
sort(unique(nhl_edge_clean$team))
sort(unique(mp_goalies_clean$team))
sort(unique(mp_teams_clean$team))
```

check team codes are standardized
```{r}
anti_join(
  mp_goalies_clean %>% distinct(season, team),
  nhl_edge_clean   %>% distinct(season, team),
  by = c("season", "team")
) %>% arrange(season, team) %>% print(n = 200)
```


account for shared variables 

```{r}
mp_teams_prefixed <- mp_teams_clean %>%
  rename_with(~ paste0("team_", .x), -c(season, team))


```


## Joining Data

Join Moneypuck Goalies and Team Data
```{r}
mp_master <- mp_goalies_clean %>%
  left_join(mp_teams_prefixed, by = c("season", "team"))

team_cols <- names(mp_master)[str_detect(names(mp_master), "^team_")]

mean(rowSums(is.na(mp_master[team_cols])) == length(team_cols))
```

Join Moneypuck Master Data with NHL Edge Data 
```{r}
master_all <- mp_master %>%
  left_join(nhl_edge_clean,
            by = c("season", "team", "player"),
            suffix = c("_mp", "_nhl"))

master_all %>%
  summarise(
    total_rows = n(),
    pct_has_gp = mean(!is.na(gp)),
    pct_has_sv = mean(!is.na(sv_percent)),
    pct_has_gaa = mean(!is.na(gaa))
  )
```


### Cleaning the master dataset


Games Played Threshold: 
- Goalies must have played greater than 8% of their teams total games played to be in data 
- 8% threshold accounts for the shortened 2012 (lockout), 2019 & 2020 (covid) seasons
```{r}
master_clean <- master_all %>%
  filter(gp / team_games_played >= 0.08)
```

Remove NA SV% and GAA
```{r}
master_clean <- master_clean %>%
  filter(!is.na(sv_percent),
         !is.na(gaa))
```

Remove redundant variables: 
```{r}
master_clean <- master_clean %>%
  rename(
    gp_mp  = games_played,
    gp_nhl = gp
  ) %>%
  mutate(
    toi_min = if_else(
      is.na(toi), NA_real_,
      {
        parts <- str_split(toi, ":", simplify = TRUE)
        as.numeric(parts[,1]) + as.numeric(parts[,2]) / 60
      }
    )
  )%>%
  select(-any_of(c(
    "file_id","s_c","name","position","situation",
    "team_name","team_team_1","team_position","team_situation",
    "team_penalities_for", "team_penalities_against", "toi"
  ))) 
  

```


Save Clean Dataset

```{r}
dir.create("data_clean")

write_csv(master_clean, "data_clean/master_goalie_dataset.csv")
```













