---
title: "NHL Goalies Analysis"
author: "Lana Robison"
date: "2026-02-27"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE)
```

# Research Question: How does NHL team defensive systems relate to goalie performance? 


```{r}
library(tidyverse)
library(janitor)
library(readr)
library(stringr)
library(ggplot2)
library(ggcorrplot)

master_clean <- read_csv("C:/Users/lanac/OneDrive - Baylor University/STA 43C8 Capstone/data_clean/master_goalie_dataset.csv")

```


## Basic Structure of Dataset

```{r}
dim(master_clean)
colnames(master_clean)

```

```{r}
summary(master_clean$sv_percent)
summary(master_clean$gaa)
```

- dataset contains 1,230 goalie-season observations and 151 variables


## Distributions of Key Outcomes

```{r}

ggplot(master_clean, aes(x = sv_percent)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Save Percentage (2008–2024)",
       x = "Save Percentage",
       y = "Count")

summary(master_clean$sv_percent)

```

- Save percentage is tightly clustered in this dataset, with a mean of 0.9073 and median of 0.9090
- suggests most NHL goalies perform within a narrow range, so even small differences may matter


```{r}
ggplot(master_clean, aes(x = gaa)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Goals Against Average",
       x = "GAA",
       y = "Count")

summary(master_clean$gaa)

```

- Goals against average is more spread out than save percentage


## Era Trends

```{r}
league_sv <- master_clean %>%
  group_by(season) %>%
  summarise(mean_sv = mean(sv_percent))

ggplot(league_sv, aes(x = season, y = mean_sv)) +
  geom_line() +
  geom_point() +
  labs(title = "League Average Save Percentage by Season",
       y = "Mean SV%",
       x = "Season")

```

- League-average save percentage changes over time, so goalie results should be interpreted within the season they occurred



```{r}
league_gaa <- master_clean %>%
  group_by(season) %>%
  summarise(mean_gaa = mean(gaa))

ggplot(league_gaa, aes(x = season, y = mean_gaa)) +
  geom_line() +
  geom_point() +
  labs(title = "League Average GAA by Season",
       y = "Mean GAA",
       x = "Season")
```

- League-average GAA also changes across seasons, which suggests league environment may affect traditional goalie stats

## Team Defense Distributions

```{r}
ggplot(master_clean, aes(x = team_x_goals_against)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Team Expected Goals Against")

```

- Team expected goals against varies across observations, so goalies face different defensive environments


```{r}
ggplot(master_clean, aes(x = team_high_danger_shots_against)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Team High Danger Shots Against")

```

- High-danger shots against also varies across teams, suggesting some goalies face more dangerous chances than others

```{r}
ggplot(master_clean, aes(x = team_corsi_percentage)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Team Corsi Percentage")

```

- team corsi %: measures a team's share of total shot attempts while at even strength, it is meant to determine a teams puck possession as teams who control the puck tend to generate more shot attempts

- Team Corsi percentage varies across teams, which suggests teams differ in territorial control and puck possession


## Goalie vs. Team Defense 

```{r}
ggplot(master_clean, 
       aes(x = team_x_goals_against, y = sv_percent)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  labs(title = "Save Percentage vs Team xGoals Against",
       x = "Team xGA",
       y = "SV%")


```

- plot suggests team defensive environment is related to goalie save percentage
- Goalies on teams allowing tougher defensive conditions may be more likely to post lower save percentages

```{r}
ggplot(master_clean, 
       aes(x = team_high_danger_shots_against, y = sv_percent)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm")

```

- This plot suggests that allowing more dangerous chances may be associated with worse goalie outcomes

## Workload Effects 

```{r}
ggplot(master_clean,
       aes(x = gp_nhl, y = sv_percent)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  labs(title = "Workload vs Save Percentage")
```

- Games played appears to have only a modest relationship with save percentage
- This suggests workload alone does not fully explain goalie performance

## Correlation Matrix 

```{r}
numeric_vars <- master_clean %>%
  select(
    sv_percent,
    gaa,
    gp_nhl,
    sa_60,
    team_x_goals_against,
    team_high_danger_shots_against,
    team_corsi_percentage,
    team_fenwick_percentage
  )

corr_matrix <- cor(numeric_vars, use = "complete.obs")

ggcorrplot(
  corr_matrix,
  type = "lower",
  lab = TRUE,
  hc.order = TRUE, 
  title = "Correlation Matrix: Goalie Performance & Team Defense"
)

```

- Several team defensive variables appear to be related to one another
- This suggests multicollinearity may be an issue in later regression models

## Fun Facts

```{r}
best_season_sv <- master_clean %>%
  filter(gp_nhl >= 50) %>%
  arrange(desc(sv_percent)) %>%
  select(player, season, team, sv_percent, gp_nhl) %>%
  slice(1:20)

best_season_sv

```

- Several elite goalie-seasons stand out in the sample, including Tim Thomas in 2010 and Igor Shesterkin in 2021


#### Goalie Instability: Count Unique Goalies per Team-Season

```{r}
goalie_counts <- master_clean %>%
  group_by(season, team) %>%
  summarise(
    n_goalies = n_distinct(player),
    .groups = "drop"
  ) %>%
  arrange(desc(n_goalies))

goalie_counts %>% slice(1:15)

```

- Some teams used many different goalies in a single season, suggesting instability in net

#### Goalie Instability: Teams with Most Goalie Turnover Overall (2008 - 2024)

```{r}
team_turnover <- master_clean %>%
  group_by(team) %>%
  summarise(
    total_goalies = n_distinct(player),
    seasons = n_distinct(season),
    avg_goalies_per_season = n() / seasons,
    .groups = "drop"
  ) %>%
  arrange(desc(avg_goalies_per_season))

team_turnover

```

