---
title: "Stat Output"
author: "Lana Robison"
date: "2026-04-01"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE)
```

# Research Question: How do NHL team defensive systems influence traditional goalie statistics? 

## Load Packages and Data 
```{r}
library(tidyverse)
library(tidymodels)
library(broom)
library(GGally)
library(car)
library(ggcorrplot)


nhl_dat <- read_csv("C:/Users/lanac/OneDrive - Baylor University/STA 43C8 Capstone/data_clean/master_goalie_dataset.csv")

```


## Define Response Variable and Build Broad Predictor Variable Pool

- Save percentage was chosen as the response variable because it is a direct measure of goalie performance
- builds a broad predictor pool from dataset
- Team totals were converted to per-game rates to make defensive context more comparable across teams and seasons
- Variables closely tied to goals allowed or save outcomes were excluded to reduce leakage into the response


```{r}
response_var = "sv_percent"

# Create rate-based defensive/context predictors
goalie_model_df = nhl_dat |>
  mutate(
    games_scale = if_else(team_games_played > 0, team_games_played, NA_real_),

    # Team strength / territorial control
    team_goals_for_pg = team_goals_for / games_scale,

    # Defensive environment - volume / shot quality / play outcomes
    team_x_goals_against_pg = team_x_goals_against / games_scale,
    team_shots_on_goal_against_pg = team_shots_on_goal_against / games_scale,
    team_shot_attempts_against_pg = team_shot_attempts_against / games_scale,
    team_unblocked_shot_attempts_against_pg = team_unblocked_shot_attempts_against / games_scale,
    team_blocked_shot_attempts_against_pg = team_blocked_shot_attempts_against / games_scale,
    team_rebounds_against_pg = team_rebounds_against / games_scale,
    team_freeze_against_pg = team_freeze_against / games_scale,
    team_play_stopped_against_pg = team_play_stopped_against / games_scale,
    team_play_continued_in_zone_against_pg = team_play_continued_in_zone_against / games_scale,
    team_play_continued_outside_zone_against_pg = team_play_continued_outside_zone_against / games_scale,

    # Turnover / discipline / support context
    team_takeaways_against_pg = team_takeaways_against / games_scale,
    team_giveaways_against_pg = team_giveaways_against / games_scale,
    team_d_zone_giveaways_against_pg = team_d_zone_giveaways_against / games_scale,
    team_penalties_against_pg = team_penalties_against / games_scale,
    team_penality_minutes_against_pg = team_penality_minutes_against / games_scale,

    # Danger-based team defense
    team_low_danger_shots_against_pg = team_low_danger_shots_against / games_scale,
    team_medium_danger_shots_against_pg = team_medium_danger_shots_against / games_scale,
    team_high_danger_shots_against_pg = team_high_danger_shots_against / games_scale,
    team_low_dangerx_goals_against_pg = team_low_dangerx_goals_against / games_scale,
    team_medium_dangerx_goals_against_pg = team_medium_dangerx_goals_against / games_scale,
    team_high_dangerx_goals_against_pg = team_high_dangerx_goals_against / games_scale,

    # Shot credit / adjusted pressure
    team_total_shot_credit_against_pg = team_total_shot_credit_against / games_scale,
    team_score_adjusted_total_shot_credit_against_pg = team_score_adjusted_total_shot_credit_against / games_scale,
    team_score_flurry_adjusted_total_shot_credit_against_pg = team_score_flurry_adjusted_total_shot_credit_against / games_scale
  )

candidate_vars = c(
  # team strength / support
  "team_x_goals_percentage",
  "team_corsi_percentage",
  "team_fenwick_percentage",
  "team_goals_for_pg",

  # defensive environment
  "team_x_goals_against_pg",
  "team_shots_on_goal_against_pg",
  "team_shot_attempts_against_pg",
  "team_unblocked_shot_attempts_against_pg",
  "team_blocked_shot_attempts_against_pg",
  "team_rebounds_against_pg",
  "team_freeze_against_pg",
  "team_play_stopped_against_pg",
  "team_play_continued_in_zone_against_pg",
  "team_play_continued_outside_zone_against_pg",

  # turnover / discipline
  "team_takeaways_against_pg",
  "team_giveaways_against_pg",
  "team_d_zone_giveaways_against_pg",
  "team_penalties_against_pg",
  "team_penality_minutes_against_pg",

  # danger profile
  "team_low_danger_shots_against_pg",
  "team_medium_danger_shots_against_pg",
  "team_high_danger_shots_against_pg",
  "team_low_dangerx_goals_against_pg",
  "team_medium_dangerx_goals_against_pg",
  "team_high_dangerx_goals_against_pg",

  # adjusted pressure metrics
  "team_total_shot_credit_against_pg",
  "team_score_adjusted_total_shot_credit_against_pg",
  "team_score_flurry_adjusted_total_shot_credit_against_pg",

  # goalie context
  "gp_nhl",
  "gs",
  "toi_min"
)

model_data = goalie_model_df |>
  select(player, season, team, all_of(response_var), all_of(candidate_vars)) |>
  drop_na()


```


## Create train/ test split

```{r}
set.seed(123)

data_split = initial_split(model_data, prop = 0.8)
train_data = training(data_split)
test_data = testing(data_split)

```


## Create Cross-Validation Folds

```{r}
cv_folds = vfold_cv(train_data, v = 5)

```


## Create Common Preprocessing recipe

- standardizes predictors and marks player, season, and team as IDS, not predictors

```{r}
base_recipe = recipe(
  as.formula(paste(response_var, "~ .")),
  data = train_data
) |>
  update_role(player, season, team, new_role = "id variable") |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

```


## Run Lasso for Variable Selection 

- Lasso was used to reduce the broad predictor pool in a data-driven way
- The lasso model selected a penalty of 0.0000187
- Lasso retained 21 predictors, suggesting goalie save percentage may be associated with several aspects of team defensive environment.

```{r}

lasso_model = linear_reg(mixture = 1, penalty = tune()) |>
  set_engine("glmnet")

lasso_wf = workflow() |>
  add_recipe(base_recipe) |>
  add_model(lasso_model)

penalty_grid = grid_regular(
  penalty(range = c(-6, 1)),
  levels = 100
)

lasso_tune = tune_grid(
  lasso_wf,
  resamples = cv_folds,
  grid = penalty_grid,
  metrics = metric_set(rmse, rsq, mae)
)

collect_metrics(lasso_tune)
autoplot(lasso_tune)

best_lasso = lasso_tune |>
  select_best(metric = "rmse")

best_lasso

final_lasso_wf = finalize_workflow(lasso_wf, best_lasso)

final_lasso_fit = final_lasso_wf |>
  fit(data = train_data)

lasso_coefs = final_lasso_fit |>
  extract_fit_parsnip() |>
  tidy()

lasso_coefs

selected_vars = lasso_coefs |>
  filter(term != "(Intercept)", estimate != 0) |>
  pull(term)

selected_vars

```

- retained variables were treated as the strongest candidates for the regression model in the original variable space


## Run PCA for multicollinearity

- uses PCA to study the correlation structure and prepare for PCR
- helps determine whether multicollinearity is severe enough to justify a component-based model
- PCA suggests that many defensive variables measure overlapping parts of team environment
- This supports using a component-based model to address multicollinearity

```{r}
# Correlation matrix first
train_predictors_only = train_data |>
  select(all_of(candidate_vars))

cor_mat = cor(train_predictors_only, use = "pairwise.complete.obs")

ggcorrplot(
  cor_mat,
  hc.order = TRUE,
  type = "lower",
  lab = FALSE
)

# PCA recipe for exploratory variance summary
pca_recipe = recipe(
  as.formula(paste(response_var, "~ .")),
  data = train_data
) |>
  update_role(player, season, team, new_role = "id variable") |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_predictors(), threshold = 0.95)

pca_prep = pca_recipe |>
  prep()

pca_variance = tidy(pca_prep, number = 3, type = "variance")
pca_loadings = tidy(pca_prep, number = 3)

pca_variance
pca_loadings


pca_scores = bake(pca_prep, new_data = NULL)

pc_names = names(pca_scores) |>
  keep(~ str_detect(.x, "^PC")) |>
  head(4)

ggpairs(
  pca_scores |>
    select(all_of(pc_names))
)

```



## Fit lasso model

- fits an interpretable regression model using the lasso-selected variables

```{r}
lasso_formula = as.formula(
  paste(response_var, "~", paste(selected_vars, collapse = " + "))
)

lasso_recipe = recipe(lasso_formula, data = train_data)

model = linear_reg() |>
  set_engine("lm")

lasso_wf = workflow() |>
  add_recipe(lasso_recipe) |>
  add_model(model)

lasso_fit = lasso_wf |>
  fit(data = train_data)

lasso_fit |>
  extract_fit_engine() |>
  tidy(conf.int = TRUE)

lasso_fit |>
  extract_fit_engine() |>
  glance()

lasso_fit |>
  extract_fit_engine() |>
  vif()

```

- Several team defensive variables are statistically associated with save percentage in this model
- The lasso-selected regression has $R^2 = 0.472$ and adjusted $R^2 = 0.460$, so it explains a moderate amount of variation in save percentage
- VIF values are very large for several predictors, indicating substantial multicollinearity, because of this, coefficient-level interpretation should be made cautiously


## Fit principal component regression model

- fits PCR as the multicollinearity-focused alternative

```{r}
# Tune the number of PCs 
pcr_recipe = recipe(
  as.formula(paste(response_var, "~ .")),
  data = train_data
) |>
  update_role(player, season, team, new_role = "id variable") |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_predictors(), num_comp = tune())

pcr_wf = workflow() |>
  add_recipe(pcr_recipe) |>
  add_model(model)

pcr_grid = tibble(num_comp = 1:length(candidate_vars))

set.seed(123)

pcr_tune = tune_grid(
  pcr_wf,
  resamples = cv_folds,
  grid = pcr_grid,
  metrics = metric_set(rmse, rsq, mae)
)

collect_metrics(pcr_tune)
autoplot(pcr_tune)

best_pcr = pcr_tune |>
  select_best(metric = "rmse")

best_pcr

final_pcr_wf = finalize_workflow(pcr_wf, best_pcr)

pcr_fit = final_pcr_wf |>
  fit(data = train_data)

pcr_fit |>
  extract_fit_engine() |>
  tidy(conf.int = TRUE)

pcr_fit |>
  extract_fit_engine() |>
  glance()

pcr_fit |>
  extract_fit_engine() |>
  vif()

```

- PCR selected 27 principal components
- The PCR model has $R^2 = 0.473$ and adjusted $R^2 = 0.457$, which is very similar to the lasso-selected regression
- All PCR VIF values equal 1, so multicollinearity was removed in this specification
- PCR is more stable statistically, but less directly interpretable in hockey terms


## Diagnostics for lasso-informed regression model

- check regression assumptions
- plots were used to check linearity, constant variance, normality, and potential influence issues
- helps determine whether the regression results are appropriate for interpretation 

```{r}
lasso_diag = lasso_fit |>
  extract_fit_engine() |>
  augment()

ggplot(lasso_diag, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Lasso: Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  )

ggplot(lasso_diag, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Lasso: Normal Q-Q Plot")

ggplot(lasso_diag, aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess", color = "blue") +
  labs(
    title = "Lasso: Scale-Location Plot",
    x = "Fitted values",
    y = "Sqrt(|Standardized Residuals|)"
  )

ggplot(lasso_diag, aes(x = .hat, y = .std.resid)) +
  geom_point() +
  labs(
    title = "Lasso: Leverage vs Standardized Residuals",
    x = "Leverage",
    y = "Standardized residuals"
  )

```

- checks were included so model quality could be assessed beyond predictive performance alone


## Diagnostics for PCR and fit ridge

- PCR diagnostics were also reviewed to assess whether the model fit was reasonable
- Ridge regression was included as an additional benchmark model


```{r}

pcr_diag = pcr_fit |>
  extract_fit_engine() |>
  augment()

ggplot(pcr_diag, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "PCR: Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  )

ggplot(pcr_diag, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "PCR: Normal Q-Q Plot")

ridge_model = linear_reg(mixture = 0, penalty = tune()) |>
  set_engine("glmnet")

ridge_wf = workflow() |>
  add_recipe(base_recipe) |>
  add_model(ridge_model)

set.seed(123)

ridge_tune = tune_grid(
  ridge_wf,
  resamples = cv_folds,
  grid = penalty_grid,
  metrics = metric_set(rmse, rsq, mae)
)

collect_metrics(ridge_tune)
autoplot(ridge_tune)

best_ridge = ridge_tune |>
  select_best(metric = "rmse")

best_ridge

final_ridge_wf = finalize_workflow(ridge_wf, best_ridge)

final_ridge_fit = final_ridge_wf |>
  fit(data = train_data)

final_ridge_fit |>
  extract_fit_parsnip() |>
  tidy()

```




## Compare models on the test set

- compares everything on the same test set

```{r}
lasso_penalized_pred = predict(final_lasso_fit, new_data = test_data) |>
  bind_cols(test_data |>
              select(player, season, team, all_of(response_var))) |>
  rename(actual = all_of(response_var), predicted = .pred) |>
  mutate(model = "lasso_penalized")

lasso_regression_pred = predict(lasso_fit, new_data = test_data) |>
  bind_cols(test_data |>
              select(player, season, team, all_of(response_var))) |>
  rename(actual = all_of(response_var), predicted = .pred) |>
  mutate(model = "lasso_selected_regression")

pcr_pred = predict(pcr_fit, new_data = test_data) |>
  bind_cols(test_data |>
              select(player, season, team, all_of(response_var))) |>
  rename(actual = all_of(response_var), predicted = .pred) |>
  mutate(model = "pcr")

ridge_pred = predict(final_ridge_fit, new_data = test_data) |>
  bind_cols(test_data |>
              select(player, season, team, all_of(response_var))) |>
  rename(actual = all_of(response_var), predicted = .pred) |>
  mutate(model = "ridge")

all_preds = bind_rows(
  lasso_penalized_pred,
  lasso_regression_pred,
  pcr_pred,
  ridge_pred
)

model_summary = all_preds |>
  group_by(model) |>
  summarise(
    rmse = rmse_vec(actual, predicted),
    rsq = rsq_vec(actual, predicted),
    mae = mae_vec(actual, predicted)
  ) |>
  arrange(rmse)

model_summary

```

- PCR, lasso-selected regression, and penalized lasso performed very similarly on the test set
- Ridge performed somewhat worse than the other models
- Overall, team defensive context helps explain save percentage, but no single model overwhelmingly outperformed the others


## Example: Real Goalie

- translates the model into a practical hockey application
- goal is to show how predicted save percentage changes under different defensive environments
- helps connect the statistical model back to player evaluation

```{r}
example_goalie = model_data |>
  filter(player == "Jonathan Quick", season == 2020) |>
  slice(1)

example_observed = example_goalie |>
  mutate(scenario = "Observed environment")

example_better_defense = example_goalie |>
  mutate(
    team_high_danger_shots_against_pg = team_high_danger_shots_against_pg * 0.90,
    team_x_goals_against_pg = team_x_goals_against_pg * 0.95,
    team_rebounds_against_pg = team_rebounds_against_pg * 0.90,
    scenario = "Improved team defense"
  )

example_compare = bind_rows(
  example_observed,
  example_better_defense
)

example_compare |>
  mutate(
    predicted_sv_lasso = predict(lasso_fit, new_data = example_compare)$.pred,
    predicted_sv_pcr = predict(pcr_fit, new_data = example_compare)$.pred
  ) |>
  select(player, season, team, scenario, sv_percent, predicted_sv_lasso, predicted_sv_pcr)

```









